Performance of Individual Audit Filters in Predicting Opportunities for Improvement

### 2024-06-16 manuscript.Rmd
update

Trauma, defined as a physical injury and the body's associated response,
accounts for 8-9% of global deaths annually [@world2014injuries;
@roth2018global], mainly affecting younger individuals in low- and
middle-income countries attributed to mainly road traffic injuries
[@vos2020global; @roth2018global; @SweTrau2022Arsrapport]. Furthermore,
trauma is responsible for the top cause for disability-adjusted
life-years (DALY) [@vos2020global] and a higher propensity towards
mental disorders and cognitive decline [@kessler1995posttraumatic;
@north1999psychological; @muscatelli2017prevalence; @newnam2014impacts;
@visser2017course]. This negatively affects individuals, communities and
society as a whole [@newnam2014impacts].It is estimated that the cost of
road traffic injuries alone accounts for 2% of gross domestic product
(GDP) in high-income countries [@world2014injuries].

@@

## Setting

```{=html}
<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->
```
## Participants

```{=html}
<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->
```
```{=html}
<!-- Case-control study: Give the eligibility criteria, and the
sources and + methods of case ascertainment and control
selection. Give the rationale for the choice of cases and
controls. For matched studies, give matching criteria and the number
of controls per case -->
```
```{=html}
<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->
```
## Variables and data sources/measurements

```{=html}
<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->
```
## Bias

<!-- Describe any efforts to address potential sources of bias -->

@@

```{=html}
<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->
```
## Statistical methods

```{=html}
<!--

@@

Figure 1: Flowchart of the patient selection process. Shows the patients
excluded and included in this study. <!--

@@

##Interpretation

Higher ISS, longer times to CT, longer times to first intervention and
higher levels of care were correlated with OFIs, corresponding with a
recent study conducted at Karolinska University Hospital by Albaaj et
al. [@albaaj2023patient]. Notably, GCS at ED showed a correlation with OFI even though the median
GCS and its IQR was identical for individuals with OFI and without. Upon
further analysis this could be explained by the difference in
distribution outside the IQR.

The AF "GCS\<9 and not intubated" showed no significant ability to
predict OFIs, similarly shown by Willis et al. [@willis2008interpreting]
in 2008. An explanation for this could be that an abnormal GCS as a
result from intoxication versus head injury for example are not
comparable, suggesting that the AF might be more reflective of clinical
decision rather than quality of care [@willis2008interpreting]. However,
according to Sanddal et al. [@sanddal2011analysis] airway management was
associated with OFI, suggesting that an alternative AF related to airway
management might perform differently.

Both AFs measuring delay: "\>60 until first intervention" and "\>30min
until first CT", showed a correlation with OFI. Comparably, Teixeira et al. [@teixeira2007preventable] found a correlation between delay
and preventable death. Furthermore, Ghorbani et al.
[@ghorbani2018analysis] identified delay in treatment and delay to CT as
common errors in trauma care. A more recent study in Japan by Yamamoto
et. al [@yamamoto2023immediate] showed a significant reduction in
mortality when patients received a whole-body CT within 10 minutes of
arrival at the ED. Notably, OFI in this study was not soley a reflection
of mortality but also morbidity. Rather, it could highlight differences in quality gaps
between hospitals.

@@

AFs, although widely used, there is lack
of consensus on what they are, what purpose they serve and how they
shall be used [@stelfox2010quality; @stelfox2011evidence;
@berg2022perceived]. Further, AFs are static in their criteria, contrary to health care in general.
Consequently, the utility of AFs may diminish with emerging challenges
or upon resolution of the quality gaps they were designed for. This
highlights the need for consistent and evidence-based review to
determine current usefulness [@santana2014development]. However, such
review process might present with costs greater than the potential
benefit from AFs.

While AFs may help determining OFIs, they could overshadow other factors
contributing to that analysis, resulting in over-reliance. AFs might
disproportionally affect certain groups of people failing to recognise
patient cases that fall outside the scope of AF criteria
[@willis2008interpreting]. For example, AFs primarily focus on aspects
of trauma care mostly relevant to medical doctors despite the
interdisciplinary nature of trauma care. This is especially
counterintuitive given the broad definition of OFIs. Lastly, confounding
variables adds to the uncertainty of AFs performance. For example,
patients receiving the same interventions at different time points may
do so as a function of injury severity not quality of care
[@willis2008interpreting].

Due to the
study being conducted in a single, trauma level I centre in Stockholm it
is unrealistic to assume applicability of these findings at other
hospitals facing unique challenges [@world2009guidelines]. It is previously known that
low and middle income countries seem to benefit from certain AFs that
high income countries do not and vice versa [@berg2022perceived].

@@

Given the expanding field of artificial intelligence and machine
learning, newer studies should focus on using these tools to screen for
patient cases at risk of an OFI. This will diminish misclassification
rates and ensure sustained high performance, attributable
to an algorithm's capability to adapt and learn from the prevailing
context of trauma care. This approach is contemporary but not
unexplored. A recent preprint by Attergrim et al.
[@attergrim2023predicting] explored the idea of machine learning models
as a screening tool for potential OFIs. They found that machine learning
models outperformed the AFs currently used at Karolinska University
Hospital in predicting OFIs.

Should the use of AFs remain, it is important to ensure its performance
by routinely updating them in order to better reflect the current
quality gaps. 
<!--

@@

>>>>

Trauma, defined as a physical injury and the body's associated response,
causes 8-9% of global deaths annually [@world2014injuries;
@roth2018global], disproportionately affecting younger individuals in low- and
middle-income countries, primarily due to road traffic injuries
[@vos2020global; @roth2018global; @SweTrau2022Arsrapport]. Trauma also ranks highest in disability-adjusted life-years (DALYs) [@vos2020global] and increases the risk of mental disorders and cognitive decline [@kessler1995posttraumatic;
@north1999psychological; @muscatelli2017prevalence; @newnam2014impacts;
@visser2017course], impacting individuals, communities and
society at large [@newnam2014impacts]. Road traffic injuries alone cost approximately 2% of gross domestic product
(GDP) in high-income countries [@world2014injuries].

@@

## Setting
Karolinska University Hospital, a Trauma level I
hospital, treats around 1,500 trauma patients annually. If a patient case results in team activation or the
patient has an Injury Severity Score (ISS) of less than 9
retrospectively, it is added to the Karolinska
trauma registry which reports to SweTrau. The registry includes data on vital
signs, times, injuries, and interventions as well as patient
demographics according to the European Consensus Statement, the Utstein
template [@ringdal2008utstein]

The Karolinska trauma registry also contains a local care quality
database with specific AFs and OFIs that are determined in MMC
conferences through consensus decisions. OFIs are assessed in multiple stages,
characterised by escalating levels of scrutiny. Notably, instances of
mortality are directly referred to the multidisciplinary conference,
where, in addition to assessing OFIs, a decision is made regarding
whether the death was preventable or potentially preventable, a
classification also falling under the purview of OFIs.

From 2013 to 2017, efforts focused on identifying adverse outcomes unrelated to mortality. The review process was refined and formalized, with trauma cases initially assessed by a specialized trauma nurse for potential OFIs. In 2017, this procedure became standard practice, incorporating preliminary evaluations by specialized trauma nurses. Trauma quality data screened for AFs and cases flagged for possible care failures underwent secondary review by two specialized nurses. Potential OFIs identified in this review prompted a comprehensive evaluation of the trauma patient's case.
```{=html}
<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->
```
## Participants

The study included 8,309 patients treated at Karolinska University Hospital from 2012 to 2022 who underwent OFI screening. Patients under 15 and those not screened for OFIs were excluded due to differences in clinical management for minors.

```{=html}
<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->
```
```{=html}
<!-- Case-control study: Give the eligibility criteria, and the
sources and + methods of case ascertainment and control
selection. Give the rationale for the choice of cases and
controls. For matched studies, give matching criteria and the number
of controls per case -->
```
```{=html}
<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->
```
## Variables and data sources/measurements
The outcome variable in this study is an OFI, as established by the MMCs
and furthermore treated as a dichotomous variable with "Yes - At least
one OFI identified" and "No - No OFI identified".
The 11 AFs used at Karolinska University Hospital served as the exposure variable, as shown in Table 1. Each AF is labeled "original" or "manually created." Manually created AFs were calculated using trauma registry data, while original AFs were recorded in the quality database based on criteria beyond the available registry data.

The data spanning 2012 to 2022 was retrieved from the
Karolinska University Hospital trauma registry and trauma care quality
database. Patient data on vital signs, care processes and interventions, level of
care and time aspects was retrieved from SweTrau while both exposures
(AFs) and outcome (OFI) were retrieved from the Karolinska University
Hospital trauma care quality database. All data was anonymised to protect patient privacy.
```{=html}
<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->
```
## Bias
Consensus decisions at MMC conferences may introduce misclassification bias. Additionally, AF flagging by specialized nurses is a manual process, potentially leading to selection bias. Healthcare professionals attending MMCs may not accurately recall key details impacting OFI decisions, introducing recall bias. Furthermore, professionals not present for specific cases might over- or underestimate OFI significance.
<!-- Describe any efforts to address potential sources of bias -->

@@

```{=html}
<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->
```
## Statistical methods
The study utilized R [@R2022language] for statistical analysis. Patient characteristics between cases with and without OFI were compared using the Wilcoxon signed-rank test for continuous variables and the Chi-squared test for categorical variables.

The predictive performance of AFs was assessed by calculating sensitivity and specificity for each AF in predicting OFI. Receiver operating characteristic (ROC) curves were constructed for binary classification, and the area under the curve (AUC) was computed to compare AF performance. [@robin2011proc].

Bootstrap resampling (1,000 iterations) was applied to estimate 95% confidence intervals for sensitivity, specificity, and AUC. The significance of each AF's AUC was evaluated against chance (AUC = 0.5) and the minimum standard for diagnostic tests (AUC = 0.8) [@nahm2022receiver] using an empirical bootstrap test (1,000 iterations).

Missing values for AFs were treated as negative (i.e., "not flagged").
```{=html}
<!--

@@

Figure 1: Flowchart of the patient selection process. Shows the patients
excluded and included in this study.

<!--

@@

##Interpretation
Higher ISS, longer times to CT, longer times to first intervention, and higher levels of care correlated with OFIs, consistent with a recent study by Albaaj et al. [@albaaj2023patient]. Notably, GCS at the ED also showed a correlation with OFIs despite identical median GCS values, likely due to distribution differences outside the IQR.

The AF "GCS\<9 and not intubated" showed no significant ability to
predict OFIs, similarly shown by Willis et al. [@willis2008interpreting]. Possibly because the AF might be more reflective of clinical
decision rather than quality of care [@willis2008interpreting]. For instance, 
An abnormal GCS resulting from intoxication is managed differently than one caused by a head injury.

Both delay-related AFs: "\>60 until first intervention" and "\>30min
until first CT", showed a correlation with OFI. Comparably, Teixeira et al. [@teixeira2007preventable] linked delays
to preventable death and Ghorbani et al.
[@ghorbani2018analysis] identified treatment and CT delays as common errors in trauma care. A recent study in Japan by Yamamoto
et. al [@yamamoto2023immediate] found reduced mortality with whole-body CT scans within 10 minutes of ED arrival.

@@

AFs are widely used, but lack consensus on which to use, purpose, and application [@stelfox2010quality; @stelfox2011evidence;
@berg2022perceived]. They have static criteria and may not adapt to evolving healthcare challenges, or upon resolution of the quality gaps they were designed for, reducing their utility over time. This underscores the need for regular evidence-based reviews of AF performance [@santana2014development]. However, these reviews might be cost-prohibitive

While AFs can help identify an OFI, they might overshadow other contributing factors, leading to over-reliance and potentially failing to recognize cases outside their scope.
[@willis2008interpreting]. For example, AFs often focus on aspects relevant to doctors, despite the interdisciplinary nature of trauma care. Lastly, Confounding variables, such as different treatment times based on injury severity, add to the uncertainty of AFs' effectiveness.
[@willis2008interpreting].

Due to the
study being conducted in a single, trauma level I centre in Stockholm it
is unrealistic to assume applicability of these findings at other
hospitals facing unique challenges [@world2009guidelines]. For instance AFs that benefit
low and middle income may not benefit
high income countries and vice versa [@berg2022perceived].

@@

Given the expanding field of artificial intelligence and machine
learning, new studies should focus on using these tools to screen for
patient cases at risk of an OFI. This will reduce misclassification
and maintain high performance, attributable
to an algorithm's capability to adapt. A recent preprint by Attergrim et al.
[@attergrim2023predicting] found that machine learning
models outperformed the AFs currently used at Karolinska University
Hospital in predicting OFIs. Should the use of AFs remain, they must be routinely updated to reflect current
quality gaps. 
<!--

@@

Keep? However,
according to Sanddal et al. [@sanddal2011analysis] airway management was
associated with OFI, suggesting that an alternative AF related to airway
management might perform differently.

Notably, OFI in this study was not soley a reflection
of mortality but also morbidity. Rather, it could highlight differences in quality gaps
between hospitals.
